---
title: "NYC Housing Price Prediction Modeling & Analysis"
output: html_notebook
---

Install libraries

```{r}
# We need the tidyverse library for data manipulation, cleaning and plotting
install.packages("tidyverse")

# We need the sf library (I'm guessing it stands for shapefile) to handle shapefiles like the Real Estate shp, dbf, shx files.
install.packages("sf")
```

Load datasets and prep

```{r}
library(tidyverse)
library(sf)

set.seed(67)

# 1. Load SLD Data (csv)
sld_data <- read_csv("sld_ny_city_with_zips.csv") 

# 2. Load Real Estate Data (shapefile)
real_estate_data_2017 <- st_read("real_property_sales_nyc_2017.shp")
real_estate_data_2018 <- st_read("real_property_sales_nyc_2018.shp")
```
Merging SLD dataset and real_estate dataset

```{r}
# Standardize Zip Codes ("zip" column is lowercase in real_estate and uppercase in SLD)
real_estate_data_2017 <- mutate(real_estate_data_2017, zip = as.numeric(zip))
real_estate_data_2018 <- mutate(real_estate_data_2018, zip = as.numeric(zip))
sld_data <- mutate(sld_data, ZIP = as.numeric(ZIP))

# Combine the 2017 and 2018 Real Estate datasets
real_estate_data <- bind_rows(real_estate_data_2017, real_estate_data_2018)

# We aggregate the SLD data by Zip Code to avoid data explosion due to ZIP not being specific enough to describe each census block group
# This compresses the detailed Census Block data into Zip Code averages
sld_grouped <- group_by(sld_data, ZIP)

sld_by_zip <- summarise(sld_grouped,
                        
    # --- DENSITY (D1) ---
    D1A = mean(D1A, na.rm = TRUE),          # Gross Residential Density (Housing Units per Acre)
    D1B = mean(D1B, na.rm = TRUE),          # Gross Population Density (People per Acre)
    D1C = mean(D1C, na.rm = TRUE),          # Gross Employment Density (Jobs per Acre)
    
    # --- DIVERSITY (D2 - Land Use Mix) ---
    D2A_JPHH = mean(D2A_JPHH, na.rm = TRUE),     # Jobs per Household (Balance of Commercial vs Residential)
    D2B_E8MIX = mean(D2B_E8MIX, na.rm = TRUE),   # Employment Mix (Diversity of job types available)
    D2R_JOBPOP = mean(D2R_JOBPOP, na.rm = TRUE), # Job to Population Balance
    D2A_EPHHM = mean(D2A_EPHHM, na.rm = TRUE),   # Employment & Household Entropy (Mixed-Use Score 0-1)
    
    # --- DESIGN (D3 - Walkability Proxies) ---
    D3A = mean(D3A, na.rm = TRUE),          # Total Road Network Density (Miles of road per sq mile)
    D3B = mean(D3B, na.rm = TRUE),          # Street Intersection Density (Higher = More walkable grid)
    
    # --- TRANSIT ACCESSIBILITY (D4) ---
    D4A = mean(D4A, na.rm = TRUE),          # Distance to nearest transit stop (meters) - Lower is better
    D4B025 = mean(D4B025, na.rm = TRUE),    # Proportion of jobs within 0.25 miles of transit
    D4D = mean(D4D, na.rm = TRUE),          # Transit Frequency (How often bus/subway comes per hour)
    
    # --- DESTINATION ACCESSIBILITY (D5) ---
    D5AR = mean(D5AR, na.rm = TRUE),        # Regional Accessibility to Jobs via Auto (Centrality)
    D5AE = mean(D5AE, na.rm = TRUE),        # Regional Accessibility to Working Population via Auto
    
    # --- WALKABILITY & LIFESTYLE ---
    NatWalkInd = mean(NatWalkInd, na.rm = TRUE), # National Walkability Index (Score 1-20)
    
    # --- LUXURY / BUSINESS PROXIES ---
    R_HiWageWk = mean(R_HiWageWk, na.rm = TRUE), # Ratio of High Wage Workers (Neighborhood Wealth Proxy)
    D1C5_OFF = mean(D1C5_OFF, na.rm = TRUE),     # Office Job Density (Indicates Business District/Downtown)
    D1C5_ENT = mean(D1C5_ENT, na.rm = TRUE),     # Entertainment Job Density (Indicates Nightlife/Trendy Area)
    
    # --- RAW COUNTS ---
    TotPop = sum(TotPop, na.rm = TRUE),     # Total Population in Zip Code
    TotEmp = sum(TotEmp, na.rm = TRUE),     # Total Employment in Zip Code
    
    # --- OTHERS ---
    Pct_AO0 = mean(Pct_AO0, na.rm = TRUE)        # Percent of Households with ZERO cars (High in NYC core)
  )

# Merge Real Estate Data (Left) with Aggregated SLD Data (Right)
housing_merged <- left_join(real_estate_data, sld_by_zip, by = c("zip" = "ZIP"))

# Drop the geometry column since lm() only works with numerical values, not shapes 
clean_housing_data <- st_drop_geometry(housing_merged)
    
# Select only the columns we need for the model
housing_data_raw <- select(clean_housing_data,
    price, borough, tot_sqft, yr_built,
    bldg_ctgy, tot_unit,land_sqft,  
    
    # --- SELECTED SLD FEATURES ---
    D1A, D1B, D1C, D2A_JPHH, D2B_E8MIX, D2R_JOBPOP, D2A_EPHHM,
    D3A, D3B, D4A, D4B025, D4D, D5AR, D5AE, NatWalkInd,
    D1C5_OFF, D1C5_ENT, TotPop, TotEmp, Pct_AO0, R_HiWageWk,
  )

# Final Cleanup on Merged Dataset (since merging by Zip code is messy)
housing_data <- distinct(housing_data_raw) # Remove duplicates
housing_data <- na.omit(housing_data)      # Remove rows with missing values

housing_data <- filter(housing_data, price > 1000)        # Remove transfers ($0-$1000) (outliers)
housing_data <- filter(housing_data, price < 10000000)    # Remove luxury outliers (> $10M) (outliers)
housing_data <- filter(housing_data, tot_sqft > 100)      # Remove invalid size
housing_data <- filter(housing_data, yr_built > 0)        # Remove invalid year

bldg_counts <- group_by(housing_data, bldg_ctgy) # Group by category to count frequency
bldg_counts <- filter(bldg_counts, n() > 50)     #
housing_data <- ungroup(bldg_counts)             # Keep only categories with > 50 sales (otherwise pred won't work)

cat("\nFinal Cleaned Dataset (Real Estate merged with SLD):", nrow(housing_data), "rows\n")
```


Sanity Check
- Yields a correlation heat map to visualize correlation between price and SLD features
- This checks for whether some features in SLD are "good predictors" of price of a real estate (the target variable)
- Also checks for multicollinearity between features, this helps with feature selection

```{r}
# Select only numerical features
numeric_vars <- select(housing_data, where(is.numeric))

# Form the correlation matrix using cor()
cor_mat <- cor(numeric_vars)

# Turn into data frame
cor_melt <- as.data.frame(as.table(cor_mat))

# Plot
ggplot(cor_melt, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", limit = c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_blank()) +
  labs(title = "Correlation Matrix: Current Feature Set")
```

Regression Modelling 

```{r}
# 1. Split Data (70% Train, 15% Validation, 15% Test)
set.seed(67)

train_idx <- sample(seq_len(nrow(housing_data)), size = floor(0.7 * nrow(housing_data))) # Index at 0.7 of whole data

train_set <- housing_data[train_idx, ]     # Train set (70%)
val_test_set <- housing_data[-train_idx, ] # Validation + Test (30%)

val_idx <- sample(seq_len(nrow(val_test_set)), size = floor(0.5 * nrow(val_test_set))) # Index at 0.5 of Validation + Test set

validation_set <- val_test_set[val_idx, ] # Validation set (15%)
test_set <- val_test_set[-val_idx, ]      # Test set (15%)

cat("Training Set Rows:  ", nrow(train_set), "(70%)\n")
cat("Validation Set Rows:", nrow(validation_set), " (15%)\n")
cat("Testing Set Rows:   ", nrow(test_set), " (15%)\n")

# Create linear regression model using our current feature set on the train set
model_lm <- lm(formula = price ~ borough + tot_sqft + yr_built + bldg_ctgy + tot_unit + land_sqft + D1A + Pct_AO0 + R_HiWageWk 
               + D1B + D1C + D2A_JPHH + D2B_E8MIX + D2R_JOBPOP + D2A_EPHHM + D3A + D3B + D4A + D4B025 + D4D + D5AR + D5AE 
               + NatWalkInd + D1C5_OFF + D1C5_ENT + TotEmp, 
               data = train_set)

summary(model_lm)
```

Evaluation (RMSE)

```{r}
# Predict using our model on the test set
preds <- predict(model_lm, newdata = test_set)

# Calculate RMSE
mse <- mean((test_set$price - preds)^2)
rmse <- sqrt(mse)
cat("Model RMSE: $", format(round(rmse, 0), big.mark = ","), "\n")

# --- VISUALIZATION: Predicted vs Actual ---
plot_data <- data.frame(Actual = test_set$price, Predicted = preds)

ggplot(plot_data, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", size = 1) +
  labs(title = "Diagnostic: Actual vs. Predicted Prices",
       subtitle = "Points near the red line are accurate predictions",
       x = "Actual Price ($)", 
       y = "Predicted Price ($)") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()
```

Cross-Validation (Manual 10-Fold)
- We want to mainly look at this to evaluate the performance of our lm model, and iteratively improve our Feature selection

```{r}
set.seed(160)
data_shuffled <- housing_data[sample(nrow(housing_data)), ]
# Get all the folds to be used in our k-fold CV
folds <- cut(seq(1, nrow(data_shuffled)), breaks = 10, labels = FALSE)

cv_mse_scores <- c()

cat("Current Feature Set:",
    "price ~ borough + tot_sqft + yr_built + bldg_ctgy + tot_unit + land_sqft + D1A + Pct_AO0 + R_HiWageWk + D1B + D1C + D2A_JPHH + D2B_E8MIX + D2R_JOBPOP + D2A_EPHHM + D3A + D3B + D4A + D4B025 + D4D + D5AR + D5AE + NatWalkInd + D1C5_OFF + D1C5_ENT + TotEmp",
    "\n")

for(i in 1:10){
  test_idx <- which(folds == i)
  test_fold <- data_shuffled[test_idx, ]
  train_fold <- data_shuffled[-test_idx, ]
  
  # Train 
  m <- lm(price ~ borough + tot_sqft + yr_built + bldg_ctgy + tot_unit + land_sqft + D1A + Pct_AO0 + R_HiWageWk + D1B + D1C + D2A_JPHH +      D2B_E8MIX + D2R_JOBPOP + D2A_EPHHM + D3A + D3B + D4A + D4B025 + D4D + D5AR + D5AE + NatWalkInd + D1C5_OFF + D1C5_ENT + TotEmp, 
          data = train_fold) # This is our best feature set so far
  
  # We remove rows from the test set that have categories the model never saw during training (e.g. category of the boulevard = 6)
  
  # Identify which columns are factors (categorical)
  factor_cols <- names(Filter(is.factor, train_fold))
  
  # Loop through each factor and filter the test set
  for(col in factor_cols) {
    # Get the levels the model learned
    train_levels <- levels(train_fold[[col]])
    
    # Filter test_fold to keep ONLY rows where the category exists in training
    # *Note that we lose a bit of data, but at least now the model works
    test_fold <- test_fold[test_fold[[col]] %in% train_levels, ]
  }
  
  # Predict
  p <- predict(m, test_fold)
  
  # Calculate Error in $ Dollars
  mse_fold <- mean((test_fold$price - p)^2)
  cv_mse_scores <- c(cv_mse_scores, mse_fold)
}

avg_mse <- mean(cv_mse_scores)
avg_rmse <- sqrt(avg_mse)

# Calculate mean price of the dataset to normalize the error (To get Error %)
mean_price <- mean(housing_data$price, na.rm = TRUE)

# Calculate Percentage Error (Cross Validation RMSE / Mean Price)
cv_rmse_pct <- avg_rmse / mean_price

# --- COMPARATIVE ANALYSIS: STRUCTURAL BASELINE ---
# We now test a model using ONLY Real Estate features from real_estate (Size, Age, Type, Land)

cat("\n--- Performance of the 3 models (CV-RMSE) ---\n")
structural_formula <- price ~ tot_sqft + yr_built + land_sqft + bldg_ctgy

# Run 10-Fold Cross-Validation on Real Estate Model
cv_mse_struct <- c()
for(i in 1:10){
  test_idx <- which(folds == i)
  test_fold <- data_shuffled[test_idx, ]
  train_fold <- data_shuffled[-test_idx, ]
  
  # Train Housing Model
  m_struct <- lm(structural_formula, data = train_fold)
  
  # Handle new levels like before
  factor_cols <- names(Filter(is.factor, train_fold))
  for(col in factor_cols) {
    if(col %in% all.vars(structural_formula)) { # Only filter if variable is in formula
      train_levels <- levels(train_fold[[col]])
      test_fold <- test_fold[test_fold[[col]] %in% train_levels, ]
    }
  }
  
  # Predict
  p <- predict(m_struct, test_fold)
  
  mse_fold <- mean((test_fold$price - p)^2)
  cv_mse_struct <- c(cv_mse_struct, mse_fold)
}

avg_rmse_struct <- sqrt(mean(cv_mse_struct))

# --- FINAL PERFORMANCE REPORT ---
baseline_errors <- housing_data$price - mean(housing_data$price)
baseline_rmse <- sqrt(mean(baseline_errors^2))
     # Evaluation Metric                         # Round to Integer                #Seperate every 3 digit with ","
cat("1. Naive Baseline (Guessing Average): $",   format(round(baseline_rmse, 0),   big.mark=","), "\n")
cat("2. Real Estate Model (Real Estate Only): $", format(round(avg_rmse_struct, 0), big.mark=","), "\n")
cat("3. Full Model (Real Estate + SLD):  $",     format(round(avg_rmse, 0),        big.mark=","), "\n")
cat("------------------------------------------------\n")
cat("Real Estate Data Lift: ", round((baseline_rmse - avg_rmse_struct)/baseline_rmse * 100, 1), "%\n")
cat("SLD Data Lift: ", round((baseline_rmse - avg_rmse)/baseline_rmse * 100, 1), "%\n")
```